### 2016-12-30 Chapters/experiments.tex
Hans modification of Material and Methods

Most studies of the growth of microorganisms have been done at steady state.
This is a reasonable and logical choice given that at steady state, the behavior of microorganisms is reproducible, which helps uncovering simple theories to apprehend their inherent complexity.
But while this condition can be easily achieved in the laboratory, microorganisms naturally spend very little time in steady state.
This motivated the construction of a new theoretical framework in Chapter~\ref{chap:theory} to study growth laws during growth transitions.

The principles underlying regulatory processes in dynamical conditions appear to be different from those applying at steady state.
We showed that regulatory strategies approaching the theoretical optimal, given by bang-bang control of gene expression machinery, have to be able to effect abrupt and strong variations in the activity of a gene.
It was seen that for implementing such strategies the regulatory system needs to be capable of sensing the internal state of the cell and not just the environment.
A near-optimal transition also requires information of several different variables, whereas for maintaining a steady state with maximal biomass accumulation a single variable turned out to be sufficient in our theoretical framework.
Interestingly, we showed that the ppGpp regulatory system of \textit{E. coli} fulfills the requirement of such a regulatory strategy, by implementing an on-off strategy for regulating the synthesis of ribosomes.
However, although this provides circumstantial evidence that bacteria control resource allocation in a manner consistent with theoretical optimality, experimental data are necessary to decide if an on-off strategy is at work in \textit{E. coli} during growth transitions.

Unfortunately, far more information is available on ribosome abundance at steady state than during growth transitions~\cite{scott_interdependence_2010,gausing_regulation_1980}.
The main reason for this bias on steady-state conditions is that, from an experimental point of view, growth transitions are hard to control and may depend on the history of the culture~\cite{ng_damage_1962,dufrenne_effect_1997,shaw_effect_1967,mcmeekin_predictive_2002,cheroutre-vialette_application_2002}.
However, as we discussed in Section~\ref{sec:discussion}, the few studies that have been reported seem to be consistent with our predictions, in the sense that they indicate that during growth transitions, the synthesis rate of ribomoses oscillates~\cite{gausing_regulation_1980,zengel_transcription_1986} and the ppGpp concentration manifests a rapid succession of increases and decreases~\cite{friesen_synthesis_1975,murray_control_2003}.
But we cannot decisively validate or disprove the model predictions from measurements that were carried out at the population level, where it is inherently hard to identify switching patterns.
What is needed for the verification of our predictions are dynamical single-cell measurements of the ribosome concentration during well-controlled growth transitions.
While the observation of an on-off strategy would show that simple models of the type introduced in Chapter~\ref{chap:theory} are instrumental in gaining a better understanding of microbial physiology, its refutation would also raise interesting questions. 
If maximization of growth rate and biomass accumulation have been retained by natural selection in \textit{E. coli}, which factors would prevent the cell from behaving optimally from a theoretical point of view?

The aim of this chapter is to measure \textit{in vivo}, during a growth transition, the resource allocation profile $\alpha (\cdot)$ and compare its dynamics with the gold standard established in Chapter~\ref{chap:theory}.
To this end, we have measured ribosomal abundance of \textit{E. coli} at the single-cell level during a nutrient upshift.
More precisely, we constructed a strain in which a fluorescent marker has been attached to a ribosomal subunit, thus allowing the \textit{in-vivo} monitoring of (changes in) the abundance of ribosomes.
In collaboration with Irina Mihalcescu of the Laboratoire Interdisciplinaire de Physique, we cultivated this \textit{E. coli} strain in a microfluidic device, allowing the long-term imaging of individual cells in well-controlled conditions, notably involving a classic upshift experiment from growth in minimal medium with acetate to minimal medium with glucose.
We then developed a Kalman smoothing method, in collaboration with Eugenio Cinquemani of Inria Grenoble -- Rh\^{o}ne-Alpes, to reconstruct $\alpha (\cdot)$  from the estimates of the variations of the growth rate and the relative ribosomal synthesis rate.
While the preliminary results presented here do not allow a decisive validation of the expected behavior, among other things due to the difficulties that were encountered during the experiments and the image analysis, we believe they are promising as a first step towards the better understanding of global resource allocation during growth transitions.

For clarity, the model of resource allocation during growth transitions developed in Chapter~\ref{chap:theory} is reproduced here, where dotted variables represent time derivatives:
\begin{eqnarray}
\dot{p}(t) &=& e_M(t)\cdot (1/\beta - r(t)) - \frac{k_R \cdot p(t)}{K_R + p(t)}\cdot r(t) \, (1+\beta\, p(t)), \label{eq:pdef-exp}\\
\dot{r}(t) &=& \frac{k_R \cdot p(t)}{K_R + p(t)}\cdot r(t) \, (\alpha(t) - \beta\, r(t)). \label{eq:rdef-exp}
\end{eqnarray}
In this form, it contains 4 variables ($e_M(t)$, $p(t)$, $r(t)$, $\alpha (t)$) and 3 parameters ($\beta$, $k_R$, $K_R$).
$e_M(t)$ [min\textsuperscript{-1}] is an indicator of the richness of the environment.
$p(t)$ [g.L\textsuperscript{-1}] is the precursor concentration inside the cell.
$r(t)$ [g.L\textsuperscript{-1}] is the concentration of gene expression machinery inside the cell.
$\alpha (t)$ [$\emptyset$] is the fraction of resources allocated to the synthesis of gene expression machinery.
$k_R$ [min\textsuperscript{-1}] is the rate constant of macromolecular synthesis.
$K_R$ [g.L\textsuperscript{-1}] is the half-saturation constant of macromolecular synthesis.
$\beta$ [L.g\textsuperscript{-1}] is the inverse of the cellular density of macromolecules, assumed to be constant.

@@

How can one measure the resource allocation variable in \textit{E. coli} cells?
There is no direct way to quantify $\alpha (\cdot)$ in an experiment, mostly because of its abstract nature.
To correctly reconstruct it, one has to know the value of every single term in the equations in which it appears.
In the model of Eqs~\ref{eq:pdef-exp}-\ref{eq:rdef-exp}, $\alpha$ only appears in Eq.~\ref{eq:rdef-exp}.
We can therefore identify $\alpha (t_k)$ for each time $t_k$ when the terms $\beta r(t_k)$, $\dot{r}(t_k)$, and $\frac{k_R \cdot p(t_k)}{K_R + p(t_k)} \cdot r(t_k)$ are known, or equivalently, the six individual components $r(t_k)$, $\dot{r}(t_k)$, $p(t_k)$, $k_R$, $K_R$, $\beta$ they are composed of.
Even when assuming that the parameters $k_R$, $K_R$, and $\beta$ are known, or at least easy to estimate independently, reconstructing the dynamics of $\alpha$ would require estimation of the (changes in) concentration of the gene expression machinery ($r$, $\dot{r}$) and the precursor concentration $p$.

Are such measurements feasible in practice?
The abundance of ribosomes can be quantified, for instance, by measuring the total RNA of the cell~\cite{scott_interdependence_2010}, or using radioactive markers \cite{gausing_regulation_1980,zengel_transcription_1986}, fluorescent labels~\cite{bakshi_superresolution_2012}, or mass spectrometry techniques~\cite{hui_quantitative_2015}.
However, only the use of fluorescent proteins complies with our need for real-time single-cell quantification.
Obtaining an estimate of precursor abundance is even more challenging, since there is no clear proxy for the totality of precursors in the cell and, while absolute quantification of all internal metabolites has been achieved in \textit{E. coli}~\cite{bennett_absolute_2009}, this method is also not suitable for a dynamical estimation in individual cells.

The problem of quantifying $\alpha (\cdot)$ can be simplified by applying a transformation of the model of Eqs~\ref{eq:pdef-exp}-\ref{eq:rdef-exp}.
Taking into account that by construction the growth rate $\mu(t)$ is given by
\[
	\mu (t) = \beta \frac{k_R \cdot p(t)}{K_R + p(t)} \cdot r(t),
\]
we can rewrite Eq.~\ref{eq:rdef-exp} as follows
\begin{equation}
\label{eq:dot_r}
\dot{r}(t) = \mu (t) \, \left(\frac{\alpha(t)}{\beta} - r(t) \right).
\end{equation}
The problem of estimating $\alpha(\cdot)$ is thus equivalent to the problem of estimating $r(\cdot)$, $\dot{r}(\cdot)$, $\mu (\cdot)$ and $\beta$.
If we are satisfied with estimating $\alpha$ up to a constant factor $\beta$, dynamically measuring the ribosome concentration and the growth rate would make $\alpha (\cdot) / \beta$ identifiable.

Given that fluorescent reporters of ribosomal proteins can provide information on both $r(\cdot)$ and $\dot{r}(\cdot)$ \cite{zulkower_robust_2015}, the above reformulation of the problem is promising for the purposes of our study.
Moreover, since the predicted bang-bang profiles in gene activity are easily hidden at the population level if cells are not synchronous, which is generally the case, expression of the ribosomal genes need to be monitored at high sampling density on the single-cell level, which is also possible with fluorescent reporters.
Inspired by previous work~\cite{bakshi_superresolution_2012}, we therefore constructed an \textit{E. coli} strain in which the S2 ribosomal subunit, encoded by the gene \textit{rpsB}, has been tagged with a green fluorescent protein (GFP).
The strain with the chromosomal \textit{rpsB-gfp} fusion produces fluorescent ribosomes that are quantifiable \textit{in vivo} in single cells while not affecting the growth rate (see Material and Methods~\ref{sec:methods_strain} and~\cite{bakshi_superresolution_2012}).
Monitoring single cells of this strain, by time-lapse fluorescence microscopy, thus makes it possible to estimate $r(\cdot)$ and $\dot{r}(\cdot)$ and reconstruct $\mu(\cdot)$.

The growth conditions also need to be carefully chosen for this study.
In natural environments, for many bacteria one of the most frequently encountered growth transitions, or at least the transition on which selection is expected to operate, is the transition from stationary phase in a nutrient-deprived medium~\cite{gefen_direct_2014} to (exponential) growth after the renewed availability of nutrients.
The transition from stationary to exponential phase is difficult to study though as many hard-to-control parameters play a role, including the duration of the nutrient stress, the composition of the medium before growth arrest, and the accumulation of waste products in the medium~\cite{ehrenberg_mediumdependent_2012}.
As a consequence, growth resumption only occurs after a lag phase of variable duration~\cite{ng_damage_1962,dufrenne_effect_1997,shaw_effect_1967,mcmeekin_predictive_2002,cheroutre-vialette_application_2002}, which is not included in the model.
For this reason, we decided to focus on a transition from exponential growth on a poor carbon source (acetate) to exponential growth on a rich carbon source (glucose) in a well-defined minimal medium.
By construction, this steady-state-to-steady-state transition requires a long acquisition time before and after the shift, respectively, to ensure that the cells have no memory of their physiological state before the transition and have enough time to reach the new steady state after the transition.
The use of the mother machine, a microfluidic device that was designed to sustain exponential growth for long periods of time, is therefore a good choice for this purpose.
It allows to maintain a constant flow of fresh medium and to effect fast transitions by switching the medium (see Material and Methods~\ref{sec:growth_condition} and \ref{sec:microflu}).
The complete experimental plan is summarized in Fig.~\ref{fig:experiment_schema}.

\begin{figure}[tb]
\centering
\includegraphics[scale=1]{./Fig/experiment_schema}
\caption{
\textbf{Schematic outline of the upshift experiment.}
The goal is to measure the fluorescence and length/area of \textit{E. coli} \textit{rpsB-gfp} cells during an acetate-to-glucose upshift.
We use M9 minimal medium supplemented with 0.2\% of acetate or glucose (see Material and Methods~\ref{sec:growth_condition}).
A preculture was started from glycerol stock for 2.5 days on 0.2\% acetate in batch condition (shake flask).
The day of the experiment, the cells were injected into the mothermachine and fed by a constant flow of fresh 0.2\% acetate medium (see Material and Methods~\ref{sec:microflu}).
Fluorescence and phase contrast images were taken every 5 minutes.
After 20~h, the feeding media was switched to 0.2\% glucose and maintained for 20~h while continuing image acquisition.
Time 0 corresponds to the moment of the acetate-glucose upshift.
}
\label{fig:experiment_schema}
\end{figure}

\subsection{Data acquisition}
\label{sec:data_acquisition}

An experiment following the plan of Fig.~\ref{fig:experiment_schema} was carried out, over a period of 40~h, but encountered a number of difficulties.
First, the motor displacing the microscope along the Z-axis intermittently stalled, thus deactivating the autofocus and the Z-compensation for the tilt of the microfluidic device in the XY plane.
As a consequence, until the experimenter manually intervened, many of the images acquired were out of focus.
This is the reason why data points between -720 and -150 min are not exploitable, but fortunately the experiment was planned in such a way that when it really mattered, notably around the growth transition, someone was present to monitor the microscope.
Second, at around 7~h after the nutrient upshift, the bacteria started to die for reasons that are unknown, the most plausible hypothesis being a phage contamination (see Discussion~\ref{sec:chap3_discussion}).
Data analysis was therefore limited to 550 min after the upshift, when roughly 3/4 of the cells were still growing.
For these and other reasons (see Section~\ref{sec:chap3_discussion}), the experiment will have to be repeated, but there was no time left for this in the framework of my PhD thesis.

The acquired data consisted of phase contrast and fluorescence images of 60 wells in total, in four different fields.
While quite a few image analysis programs have been reported in the literature~\cite{klein_tlm-tracker:_2012,wang_image_2010,young_measuring_2012,paintdakhi_oufti:_2016}, and some of these specifically address the mother machine design of the microfluidic device~\cite{wang_robust_2010,sachs_image-based_2016}, they all presented limitations when applied to our data.
For instance, the segmentation algorithms experienced difficulties due to the low resolution of the camera, and the phase contrast images had a superposed reflection band due to the microfluidic device.
Moreover, the fluorescence density was concentrated in hot spots (Fig.~\ref{fig:data_acquisition}) and its intensity varied during the experiment. 
While this was expected, since ribosomes are localized in the cell poles~\cite{bakshi_superresolution_2012} and ribosomal abundance is known to vary with the growth rate~\cite{scott_interdependence_2010}, it complicated automatic segmentation on the fluorescence images.

For these reasons, the image analysis carried out for this chapter has been much simplified. 
First, we have focused on the cell at the bottom of each well, since this avoids the tracking of individual cells across several generations and ensures that the descendance of this cell can be followed throughout the experiment.
Second, segmentation was done by manually selecting two pixels, one at each pole of the cell.
These pixels were used to create a rectangle surrounding the cell (Fig.~\ref{fig:data_acquisition} and Material and Methods~\ref{sec:cell_segmentation}).
After background correction (Material and Methods~\ref{sec:cell_segmentation}), the fluorescence intensity in units RFU was evaluated for each pixel in the rectangle, as well as the length of the cell, defined by the distance in pixels between the poles (Fig.~\ref{fig:data_acquisition}).
The fluorescence density for the entire cell [RFU/pixel/cell] was computed by dividing the sum of the fluorescence intensities of the pixels in the rectangle by the total number of pixels in the rectangle.
Although we are well aware that the above procedure can be improved on many counts (Section~\ref{sec:chap3_discussion}), we nevertheless believe that it provides a reasonable approximation of the quantities of interest and a valid starting-point for the estimation of the growth rate and the resource allocation profile in the remainder of this chapter.

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/data_acquisition}
\caption{
\textbf{Results of data acquisition.}
We imaged 6 fields (X1Y1, X2Y1, X3Y1, X1Y2, X2Y2, X3Y2) each containing 15 wells (W0 to W14) for $\sim$40 hours.
From the 90 wells, 68 were suitable for further analysis (the others being empty, out of frame for some period of time, or plugged).
For all wells, data points are missing in the interval between -720 and -150 min because the camera was out of focus.
We stopped the analysis at 550 min, when about 3/4 of the bacteria were still growing, before the entire population died within a few hours for an unknown reason.
The image labelled "Raw image" is the last image analyzed for the highlighted well.
The bacterium on the left of this image was manually segmented by selecting two pixels at the poles on the fluorescence images (red cross).
A 6-pixel-wide rectangular mask was computed for each image, resulting in the "Segmented image" on the right.
Fluorescence intensities are expressed in Relative Fluorescence Units (RFU) on a 16-bit image (values from 0 to 65335) and were corrected for camera background, but not autofluorescence background (Material and Methods~\ref{sec:cell_segmentation} and Discussion~\ref{sec:chap3_discussion}).
The fluorescence intensity of the cell, expressed in units RFU/pixel/cell, was computed by dividing the sum of the fluorescence intensities of the pixels in the rectangle by the total number of pixels in the rectangle.
The cell length is the distance in pixels between the two poles (red line).
The thick lines in green and black highlight the time-varying length and fluorescence density for the cell visible in the top images, labelled "X3Y2, W2".
The vertical dashed lines represent the time of the upshift from growth on acetate to growth on glucose.
}
\label{fig:data_acquisition}
\end{figure}

In total, we obtained time courses of fluorescence density [RFU/pixel/cell] and the length for 68 bacterial cells (Fig.~\ref{fig:data_acquisition}).
The fluorescence density appears constant during growth on acetate (before time 0) and immediately increases when the carbon source is switched to glucose.
Inspection of the cellular length shows that the division frequency abruptly increases after the nutrient upshift, corresponding to a higher growth rate.
Unfortunately, while steady-state growth on acetate was reached before the upshift, the fluorescence density profile suggests that the experiment did not continue long enough to ensure that a new steady state on glucose was reached.
However, the data before and after the nutrient upshift are exploitable.
What do we observe if we estimate the growth rate and the resource allocation profile from these data?

\subsection{Data analysis using Kalman smoothing}
\label{sec:res_kalman}

>>>>

Most studies on the growth of microorganisms are done at steady state.
It is a reasonable and logical choice given the drawbacks of working in dynamic.
At steady state, microorganism behaviors are reproducible, and which helps uncover simple theories to apprehend their inherent complexity.
But while this ideal environment is easily achievable at the bench, microorganisms naturally spend very little time in such a perfect condition.
This motivated in chapter~\ref{chap:theory} the construction of a new theoretical framework to study growth laws during growth transitions.

The principles that define the regulatory processes in dynamic appear to be different than the ones that apply at steady state.
We showed that efficient regulatory strategies have to be able to perform extremely abrupt and intense variations of the gene activity.
This relies on the fact that bang-bang synthesis of gene expression machinery settles the upper-bound for biomass production during an environmental change.
Such strategies are only implementable if the regulatory system focus on the internal state of the cell.
But not all information is equivalent, and performing a near-perfect transition might require to measure several variables, a concept totally missing from steady-state considerations.
Interestingly, we uncovered how the ppGpp regulatory system of \textit{E. coli} might be a cheap way for the cell to access just enough information, and to implement a \textit{on-off} strategy to regulate the synthesis of ribosomes.
Overall, it taught us that the actual complexity of regulatory systems is only beneficial in a dynamical context.
But elucidating what is the best behavior, and showing that \textit{E. coli} actually implement it are different matters...

Reality actually highlights the problem with adopting a dynamical perspective on growth.
There are far more information on ribosome abundance at steady state than during growth transitions.
It is widely admitted that the latter are hard to control and rely too much on hidden variables about the strain history~[citation ? Bremmer ?].
Though, as we discussed in section~\ref{sec:discussion}, some data are available and seem to confirm that during growth transitions, the  synthesis rate of ribomoses can oscillate~\cite{gausing_regulation_1980,zengel_transcription_1986}, and that the ppGpp regulatory system can go through rapid changes~\cite{friesen_synthesis_1975,murray_control_2003}.
But we cannot decisively validate or disprove the model predictions from measurements that were carried out at the population level, where it is inherently hard to identify switching patterns.
What we need are dynamical single-cell measurements of the ribosome concentration during well-controlled growth transitions.
Would the \textit{on-off} strategy be observed, it could confirm the importance of optimizing transitions for the organism.
Would it be disproved, that could raise even more interesting questions: growth rate and biomass might not necessarily be what natural selection care for, or other unexpected limitations might prevent the cell to behave optimally.

In Chapter~\ref{chap:theory}, we developed a model of resource allocation during growth transitions.
By applying optimal control theory, we predicted that in order to maximize biomass production, the cell should produce its gene expression machinery in a bang-bang manner.
The full model with dimensional parameters is reproduced below, with dotted variables representing their time derivative:
\begin{eqnarray}
\dot{p}(t) &=& e_M(t)\cdot (1/\beta - r(t)) - \frac{k_R \cdot p(t)}{K_R + p(t)}\cdot r(t) \, (1+\beta\, p(t)), \label{eq:pdef-exp}\\
\dot{r}(t) &=& \frac{k_R \cdot p(t)}{K_R + p(t)}\cdot r(t) \, (\alpha(t) - \beta\, r(t)). \label{eq:rdef-exp}
\end{eqnarray}
In this form, it contains 4 variables ($e_M(t)$, $p(t)$, $r(t)$, $\alpha (t)$) and 3 parameters ($\beta$, $k_R$, $K_R$).
$e_M(t)$ [min\textsuperscript{-1}] is the richness of the environment.
$p(t)$ [g.L\textsuperscript{-1}] is the precursor concentration inside the cell.
$r(t)$ [g.L\textsuperscript{-1}] is the gene expression machinery concentration inside the cell.
$\alpha (t)$ [$\emptyset$] is the mass ratio of resources that goes to the gene expression machinery.
$k_R$ [min\textsuperscript{-1}] is the rate constant of macromolecular synthesis.
$K_R$ [g.L\textsuperscript{-1}] is the half-saturation constant of macromolecular synthesis.
$\beta$ [L.g\textsuperscript{-1}] is the inverse of the cellular density of macromolecules (assumed constant).
The goal of this chapter is to measure in vivo $\alpha (\cdot)$ during a growth transition, and check how it compares with the gold standard established in chapter~\ref{chap:theory}.

To this end, we took the challenge of measuring ribosomal abundance of \textit{E. coli} at the single-cell level during a controlled upshift.
We made a strain displaying fluorescent ribosomes, thus allowing in-vivo measurement of their abundance in dynamical conditions.
We placed it in a microfluidic device allowing the long-term imaging of individual cells while they are presented different growing media, and focused on the classic upshift from acetate to glucose.
We then used Kalman smoothing to estimate the variations of the growth rate and the relative ribosomal synthesis rate during this upshift.
While not allowing a decisive validation of the expected behavior, this experiment opens interesting questions and is a step towards a better understanding of the gene expression machinery regulation during growth transitions.

@@

How does one measure resource allocation in real cells?
There is no direct way to measure $\alpha (\cdot)$ in an experiment, mostly because of its abstract nature.
To correctly reconstruct it, one has to know the value of every single term in the equations in which it appears.
In the full dynamical system Eqs~\ref{eq:pdef-exp}-\ref{eq:rdef-exp}, $\alpha$ only pops up in Eq.~\ref{eq:rdef-exp}, and is already isolated in this equation, which simplifies the burden.
We can identify $\alpha (t_k)$ for each time $t_k$ where the following 3 terms are known: $\beta r(t_k)$, $\dot{r}(t_k)$, and $\frac{k_R \cdot p(t_k)}{K_R + p(t_k)} \cdot r(t_k)$.
Such a consideration remains valid if one can estimate the 6 individual components $r(t_k)$, $\dot{r}(t_k)$, $p(t_k)$, $k_R$, $K_R$, $\beta$.
Even if we assume that the constant parameters are known, or at least easier to estimate independently, reconstructing the dynamics of $\alpha$ would require to co-estimate the gene expression machinery concentration ($r$, $\dot{r}$) and the precursor concentration $p$.

Are such measurements acceptable?
Measuring the abundance of gene expression is achievable if one focus on a proxy that represents the overall abundance.
Ribosomes are widely known to be the most abundant part of the gene expression machinery, which allows to consider it as its main component~\cite{scott_interdependence_2010, scott_bacterial_2011,scott_emergence_2014}.
Their abundance can be estimated for instance by using the total RNA of the cell~\cite{scott_interdependence_2010}, radioactive markers\cite{gausing_regulation_1980,zengel_transcription_1986}, or fluorescent labeling~\cite{bakshi_superresolution_2012}.
Estimating precursor abundance is more challenging.
There is no clear proxy that represents all the precursors in the cell, so using common methods would be excessively time expensive~[citation classic methods to estimate precursor ?].
While absolute quantification of all the internal metabolites has been achieved in \textit{E. coli}~\cite{bennett_absolute_2009}, this method do not comply with our need of real-time single-cell quantification.

This problem can be overcome by using a simple transformation.
Taking into account that by construction the growth rate  $\mu(t)$ is given by:
\[
	\mu (t) = \beta \frac{k_R \cdot p(t)}{K_R + p(t)} \cdot r(t),
\]
we can rewrite Eq.~\ref{eq:rdef-exp} as follows
\begin{equation}
\label{eq:dot_r}
\dot{r}(t) = \mu (t) \, \left(\frac{\alpha(t)}{\beta} - r(t) \right).
\end{equation}
The problem of estimating $\alpha(\cdot)$ is thus equivalent to the same problem where $r(\cdot)$, $\dot{r}(\cdot)$, $\mu (\cdot)$ and $\beta$ have to be co-estimated.
If we are satisfied with estimating $\alpha$ up to a constant factor $\beta$, dynamically measuring the ribosome concentration and the growth rate would make $\alpha (\cdot) / \beta$ identifiable.

These new measurements are more acceptable given the requirements of the study.
We want to be able to identify potential abrupt changes of the gene expression.
Such switching patterns are easily hidden at the population level if cells are not synchronous, which is generally the case.
A short sample time and single-cell measurements are thus necessary, which motivated the use of fluorescent reporters like the green fluorescent protein (gfp).
We constructed a strain displaying a gfp-tagged S2 ribosomal subunit (the gene name being \textit{rpsB}).
The \textit{rpsB-gfp} strain produces fluorescent ribosomes, making them quantifiable while minimizing the burden for the cell (see Material and Methods~\ref{sec:methods_strain}).
By observing such a strain through a microscope during a growth transition, we could theoretically estimate the ribosome concentration by measuring the cell fluorescence, and reconstruct the growth rate by measuring the cell size.

The growth conditions might also play a role in the limitations of the study.
In natural conditions, the most straightforward growth transitions, or at least the one at which most of the selection is expected to occur, is the transition from the stationary phase in nutrient-deprived medium to any medium sustaining growth.
There are good reasons though for which this particular growth transition is ignored from most studies.
The transition from stationary to exponential phase is dramatically ill-controlled, in the sense that a lot of hard-to-control parameters seem to play a huge role in it.
For instance, this is the case of the duration of the nutritive stress~[citation to find], the nutrient encountered during the last exponential phase~[citation to find], or the previous accumulation of wastes in the medium~[citation to find].
This somehow results in a phenomenon called the lag phase~[citation lag phase review to find], the mechanisms of which are missing from our model.
Furthermore, most microorganisms dramatically change their physiology in stationary-phase, either sporulating or organizing into biofilms~[citations subtilis sporulation and coli biofilms to find], which goes largely beyond the scope of this study.
For this reason, we decided to focus on a transition from a slow exponential growth to high exponential growth, in particular the transition from acetate to glucose.
By construction, this steady-state-to-steady-state transition requires a long acquisition time before and after the shift, respectively to ensure cells have \textit{forgotten} the previous stress phase, and have enough time to reach the new steady state.
The mothermachine, a microfluidic device that was designed to sustain exponential growth for long period of time, seemed like a good choice for this purpose.
It allows to maintain a constant flow of fresh medium around the cells, and make fast transitions possibles by switching the medium that is fed to the cells (see Material and Methods~\ref{sec:growth_condition} and \ref{sec:microflu}).
The complete experimental scheme is summarized in Fig.~\ref{fig:experiment_schema}.

\begin{figure}[tb]
\centering
\includegraphics[scale=1]{./Fig/experiment_schema.eps}
\caption{
\textbf{Scheme of the upshift experiment.}
The goal is to measure the fluorescence and growth rate of cells during an acetate-to-glucose upshift.
All the media used are M9 minimal medium supplemented with 0.2\% of the indicated carbon source, in this case acetate or glucose (see Material and Methods~\ref{sec:growth_condition}).
The \textit{rpsB-gfp} strain was precultivated for 2.5 days on 0.2\% acetate in batch condition (shaken flask).
The day of the experiment, the cells were injected into the mothermachine and fed by a constant flow of fresh 0.2\% acetate medium (see Material and Methods~\ref{sec:microflu}).
Fluorescence and phase contrast images were taken every 5 minutes.
Twenty hours later, the feeding media was switched to 0.2\% glucose (known to theoretically sustain a 5-time-higher growth rate) and was maintained for twenty more hours while imaging carried on.
Time 0 corresponds to the exact moment when cells start to be in contact with the glucose medium.
}
\label{fig:experiment_schema}
\end{figure}

\subsection{Data acquisition}

It is important to evoke most of the issues that occurred while carrying on the experiment.
At the time, a serious issue was starting to affect our only microscope: the motorization along the Z-axis was randomly stalling until an operator manually intervenes.
This deactivates the autofocus and the Z-compensation for the tilt of the device when moving along the XY plane.
In other words, series of images are out of focus and thus not exploitable if the operator was not around, which can happen a lot in a 40-hour-long experiment.
For the experiment presented here, we timed the operations so that the operator can be around as long as possible when it counts, \textit{i.e.} close to the transition at time 0.
This is the reason why data points between -720 and -150 minutes are not available (corresponding to the steady state on acetate).
Another issue, more specific to the presented occurence of the experiment, was that bacteria started to massively die several hours after the transition.
The reason is unknown, the most plausible hypothesis being a phage contamination.
\footnote{Note that the operator was present when the dying started, which excludes the implication of lab gnomes.}
Data analysis was thus stopped at 550 minutes, when roughly 3/4 of the cells were still growing.
Note that for these reasons, the results of this preliminary experiment must be taken with caution.
The necessary improvements are extensively discussed in section~\ref{sec:chap3_discussion}.

Image analysis was performed on the remaining shots.
While image analysis softwares overflow the literature~[citations schitcells + others], some of them being specifically designed for the mother machine~[molysso + Wang et al], technical limitations did not allow to use any of them.
For instance, the segmentation was systematically impeded by the poor resolution of the camera.
The phase contrast images were not directly exploitable since we noticed they were not aligned with the fluorescence images.
Fluorescence was also organized in hot spots (see Fig.~\ref{fig:data_acquisition}) and their intensity vary widely during the experiment, which was expected since ribosomes are extremely localized~\cite{bakshi_superresolution_2012} and ribosomal abundance is known to vary with the growth rate~\cite{scott_interdependence_2010}.
It, however, dramatically impeded the automatic segmentation on the fluorescence images.
In section~\ref{sec:chap3_discussion}, we give insights into how we plan to overtake these limitations.

The analysis presented in this study is thus preliminary.
First, we focused solely on the deepest cell of each well.
It avoided the tracking step, and ensured that the cell was visible from the beginning to the end of the experiment.
Secondly, the segmentation was done by manually selecting two pixels at the poles of the cell of interest.
These pixels were used to create a rectangular mask around the cell (see Fig.~\ref{fig:data_acquisition} and Material and Methods~\ref{sec:cell_segmentation}).
After background correction (see Material and Methods~\ref{sec:cell_segmentation}), the RFU concentration was evaluated in this region of interest, as well as the length of the cell, defined by the distance in pixels between the poles (see Fig.~\ref{fig:data_acquisition}).

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/data_acquisition}
\caption{
\textbf{Results of data acquisition.}
We imaged 6 fields (X1Y1, X2Y1, X3Y1, X1Y2, X2Y2, X3Y2) each containing 15 wells (W0 to W14) for $\sim$40 hours.
From this, only 68 wells were available for analysis (the others were either empty, out of frame for some period of time, or plugged).
For all wells, data points are missing in the time interval [-720,-150] because the camera was out of focus.
We stopped the time series analysis at 550 minutes when visually 3/4 of the bacteria were still growing.
The whole population was wiped out in the next few hours for an unknown reason.
The Raw image is the last image analyzed for the highlighted well.
The left bacteria on this image was manually segmented by selecting two pixels at the poles on the fluorescence images (red cross).
A 6-pixel-wide rectangular mask was computed for each image, giving the segmented image on the right.
Fluorescence levels are expressed in Relative Fluorescence Units (RFU) on a 16-bit image (values from 0 to 65335), and were corrected for camera background, but not autofluorescence background (see Material and Methods~\ref{sec:cell_segmentation} and Discussion~\ref{sec:chap3_discussion}).
The RFU/pixel/cell was computed by taking the ratio between the sum of the pixels in the mask, and the total number of pixels in the mask.
The bacteria length is the distance in pixels between the two poles (red line on the raw and segmented images).
The thick lines highlights the time series for the cell visible in the top images.
The dashed black line corresponds to the switch time from acetate to glucose.
}
\label{fig:data_acquisition}
\end{figure}

From this preliminary analysis, we obtained time series for the concentration of fluorescence (RFU/pixel) and the length for 68 bacterial cells (Fig.~\ref{fig:data_acquisition}).
The RFU/pixel appears constant on the acetate phase (before 0) and immediately increases when the medium is switched to glucose.
Bacterial length shows that the division time abruptly decreases after the nutrient upshift, corresponding to a higher growth rate.
Unfortunately, while the steady state on acetate was reached before the upshift, it is obvious that the time series stop before a new steady state on glucose is reached, at least for the fluorescence.
Nevertheless, the time series are not interrupted around the transition.
What do we observe if we estimate the growth rate and $\alpha$ in this region?

\subsection{Data analysis using Kalman smoother}
\label{sec:res_kalman}

As we presented in section~\ref{sec:exp_design}, our goal is to reconstruct the signal $\alpha (\cdot)$ during the transition.
By modeling \textit{E. coli} as a cylinder, the length can be assumed directly proportional to the volume of the cell.
If we assume a negligible background of autofluorescence (see Material and Methods~\ref{sec:cell_segmentation}), the fluorescence concentration in the cell can be assumed proportional to the ribosome concentration.
More precisely, based on the data available, we have the following measurement model at each time-point $t_k$:
\begin{eqnarray}
L(t_k) &= \lambda \cdot V(t_k) + \epsilon_k, \label{eq:mes_L}\\
F(t_k) &= \gamma \cdot r(t_k) + \eta_k, \label{eq:mes_F}
\end{eqnarray}
where $L(t_k)$ (resp. $F(t_k)$) is the length (resp. RFU/pixel) measured at time $t_k$ (visible on Fig.~\ref{fig:data_acquisition}), $V(t_k)$ (resp. $r(t_k)$) is the actual volume (resp. ribosome concentration) at time $t_k$, ($\lambda$,$\gamma$) are unknown proportionality constants, and ($\epsilon_k$,$\eta_k$) are measurement noises.
As we saw in section~\ref{sec:exp_design}, reconstructing $\alpha (\cdot)$ requires information on $\mu (\cdot)$, $r(\cdot)$ and $\dot{r}(\cdot)$.
From Eq.~\ref{eq:mes_F}, we could obtain $r(\cdot)$ and $\dot{r}(\cdot)$ by smoothing interpolation and differentiation.
The same can be said for Eq.~\ref{eq:mes_L} and $\mu (\cdot)$ since it is defined by:
\[
\dot{V}(t) = \mu (t) \cdot V(t).
\]
However, smoothing interpolation is particularly sensible to the boundary conditions.
Since each division in the time series is a boundary, the smoothing interpolation of this kind of data is particularly unstable.
For this reason, single-cell growth rate is usually estimated by fitting an exponential term on the volume between each division (or in particular a linear term on the logarithm of the volume)~[citation Wang et al et Izard et al].
While it is suitable when bacteria are at steady state, it is not applicable in our case where we expect growth-rate variations between divisions.
Other techniques are less sensible to this kind of problem~\cite{zulkower_robust_2015}, and we decided to focus on the use of Kalman smoother~\cite{kailath_linear_2000,jazwinski_stochastic_2007}.
